The state of knowledge for marine invertebrate biodiversity in the continental US

### 2017-06-19 manuscript.Rmd
edit manuscript

With increasing impact of human activities on the biosphere, we need efficient and accurate approaches to document changes in species ranges and community composition. While terrestrial ecosystems have received considerable attention, the baseline of healthy and resiliency of marine ecosystems are not well understood [@Worm2006], and modeling the impact of human activities is limited to a few taxonomic groups [@Tittensor2010]. A major hindrance for marine species, due to their complex life history strategies, and their size range, is that we do not have enough information about which species live where, or what portion of marine biodiversity remains to be discovered and described [@Mora2011]. Additionally, coordinated marine ecosystem monitoring initiatives are lacking [@Duffy2013,@Borja2016]. There is therefore a pressing need to improve our knowledge of marine systems before species go extinct, and their functioning is irreversibly altered.

Until very recently, our knowledge of biodiversity was scattered among heterogeneous sources (e.g. natural history collections, publications, expert knowledge), difficult to access, and therefore difficult to synthesize. The development of public databases for natural history records (e.g. iDigBio [@Page2015]), species occurrences (e.g., OBIS [@Zhang2002]), genetic data (e.g., GenBank [@Benson2013], BOLD [@Ratnasingham2007a]), and registers of species names (e.g., WoRMS [@Costello2013b]), now allow us to evaluate what and how much we know. Here, we aggregate publicly available data for marine invertebrates that occur within the limits of the exclusive economic zone (EEZ) of the continental United States, as well as the Gulf of Mexico and the Pacific Northwest, to highlight gaps in our knowledge that provide a roadmap for important new directions of global research. We show that public databases are relatively incomplete, and that most species are known from only a few specimens and a few locations. We emphasize that our understanding of marine biodiversity, and the impact of human activities on communities require not only to get a more accurate list of the species that inhabits the oceans, but we also need more comprehensive data on species distributions and abundances. This can only be achieved with additional cost-effective sampling and a commitment to training more taxonomists and experts that can identify and interpret the material collected.

@@

To compare the faunistic lists from the Gulf of Mexico and the Pacific Northwest with records in public databases, we queried the iDigBio (https://portal.idigbio.org) and OBIS (http://www.iobis.org/) databases for all records they included in the same geographic boundaries as the lists. After filtering these lists to only include the same higher taxonomic level taxa as the species lists, and keeping only records for which the species could be matched to a species name in WoRMS, we had for the Gulf of Mexico `r format(nrow(idigbio_gom_records), big.mark=",")` and `r format(nrow(obis_gom_records), big.mark=",")` records for iDigBio and OBIS respectively, representing `r format(dplyr::n_distinct(idigbio_gom_records$worms_valid_name), big.mark=",")` and `r format(dplyr::n_distinct(obis_gom_records$worms_valid_name), big.mark=",")` species. For the Pacific Northwest, we retrieved `r format(nrow(idigbio_pnw_records), big.mark=",")` and `r format(nrow(obis_kozloff_records), big.mark=",")` records for iDigBio and OBIS respectively representing `r format(dplyr::n_distinct(idigbio_pnw_records$worms_valid_name), big.mark=",")` and `r format(dplyr::n_distinct(obis_kozloff_records$worms_valid_name), big.mark=",")` species. For the Gulf of Mexico, the majority of the species included in the expert lists are recorded in both iDigBio and OBIS (`r format_output(species_in_common$n_spp_gom_idigbio_obis/species_in_common$n_spp_gom*100)` %), or only OBIS (`r format_output(species_in_common$n_spp_gom_obis/species_in_common$n_spp_gom*100)` %) (Fig. \@ref(fig:database-overlap)). For the Pacific Northwest, these proportions were much lower with only `r format_output(species_in_common$n_spp_kozloff_idigbio_obis/species_in_common$n_spp_kozloff*100)`% of the species from the regional list with records in OBIS and iDigBio, and `r format_output(species_in_common$n_spp_kozloff_obis/species_in_common$n_spp_kozloff*100)`%  in OBIS only (Fig. \@ref(fig:database-coverage)).

@@

@@

In order to predict the impact of climate change and other anthropogenic perturbations on species distributions, we need to collect more occurences for species. One potential challenge for this is to be able to convey taxonomic concepts for mostly unamed species. Developing robust identifiers that can be used across biodiversity information systems, and potentially anchored in results from DNA barcoding or other molecular techniques, is needed to collaboratively build our knowledge of species distributions.

@@

Given the urgency of the biodiversity crisis, we need to develop data-driven initiatives to efficiently document biodiversity, and build on our existing knowledge to collect information to improve our understanding of marine systems.

>>>>

With increasing impact of human activities on the biosphere, we need efficient and accurate approaches to document changes in species ranges and community composition. While terrestrial ecosystems have received considerable attention, the baseline of healthy and resilient marine ecosystems are not well understood [@Worm2006], and modeling the impact of human activities is limited to a few taxonomic groups [@Tittensor2010]. For marine organisms, because of the diversity of their complex life history strategies and their size range, we have limited information about which species live where, or what portion of their diversity remains to be discovered and described [@Mora2011]. Additionally, coordinated marine ecosystem monitoring initiatives are lacking [@Duffy2013,@Borja2016], making it difficult to monitor changes in species distributions and abundances. There is therefore a pressing need to improve our knowledge of marine systems before their functioning is irreversibly altered, and species go extinct.

The continental US includes a wide diversity of marine ecosystems from the Florida Keys, to the Gulf of Maine on the East Coast; and from the arid Southern California to the oceanic coast of the Pacific Northwest. With estuaries, bays, sandy beaches, rocky shores, and coral reefs, the US coastlines harbors many biotas. While the continental shelf extends for several hundred kilometers on the East coast, it is narrow on the West coast. This diversity of habitats and depths, combined with the exploitation of natural resources as early as the 1880, has led to extensive exploration and documentation of marine biodiversity [@Fautin2011]. However, apart from the synthesis that emerged from the Census of Marine Life efforts [@Fauntin2011], no study has attempted to synthesize the current state of knowledge of marine biodiversity in the US.

Until very recently, our knowledge of biodiversity was scattered among heterogeneous sources (e.g. natural history collections, publications, expert knowledge), difficult to access, and therefore difficult to synthesize. The development of public databases for natural history records (e.g. iDigBio [@Page2015]), species occurrences (e.g., OBIS [@Zhang2002]), genetic data (e.g., GenBank [@Benson2013], BOLD [@Ratnasingham2007a]), and registers of species names (e.g., WoRMS [@Costello2013b]), now allow us to evaluate what and how much we know. Here, we aggregate publicly available data for marine invertebrates that occur within the limits of the exclusive economic zone (EEZ) of the continental United States (US), as well as the Gulf of Mexico and the Pacific Northwest, to highlight gaps in our knowledge that provide a roadmap for important new directions of global research. We show that public databases are relatively incomplete, and that most species are known from only a few specimens and a few locations. We emphasize that our understanding of marine biodiversity, and the impact of human activities on communities require not only to get a more accurate list of the species that inhabits the oceans, but we also need more comprehensive data on species distributions and abundances. This can only be achieved with additional cost-effective sampling and a commitment to training more taxonomists and experts that can identify and interpret the material collected.

@@

To compare the faunistic lists from the Gulf of Mexico and the Pacific Northwest with records in public databases, we queried the iDigBio (https://portal.idigbio.org) and OBIS (http://www.iobis.org/) databases for all records they included in the same geographic boundaries as the lists. After filtering these lists to only include the same higher taxonomic level taxa as the species lists, and keeping only records for which the species could be matched to a species name in WoRMS, we had for the Gulf of Mexico `r format(nrow(idigbio_gom_records), big.mark=",")` and `r format(nrow(obis_gom_records), big.mark=",")` records for iDigBio and OBIS respectively, representing `r format(dplyr::n_distinct(idigbio_gom_records$worms_valid_name), big.mark=",")` and `r format(dplyr::n_distinct(obis_gom_records$worms_valid_name), big.mark=",")` species. For the Pacific Northwest, we retrieved `r format(nrow(idigbio_pnw_records), big.mark=",")` and `r format(nrow(obis_kozloff_records), big.mark=",")` records for iDigBio and OBIS respectively representing `r format(dplyr::n_distinct(idigbio_pnw_records$worms_valid_name), big.mark=",")` and `r format(dplyr::n_distinct(obis_kozloff_records$worms_valid_name), big.mark=",")` species. For the Gulf of Mexico, the majority of the species included in the expert lists are recorded in both iDigBio and OBIS (`r format_output(species_in_common$n_spp_gom_idigbio_obis/species_in_common$n_spp_gom*100)` %), or only OBIS (`r format_output(species_in_common$n_spp_gom_obis/species_in_common$n_spp_gom*100)` %) (Fig. \@ref(fig:database-overlap)). For the Pacific Northwest, these proportions were much lower with only `r format_output(species_in_common$n_spp_kozloff_idigbio_obis/species_in_common$n_spp_kozloff*100)`% of the species from the regional list with records in OBIS and iDigBio, and `r format_output(species_in_common$n_spp_kozloff_obis/species_in_common$n_spp_kozloff*100)`%  in OBIS only (Fig. \@ref(fig:database-coverage)). Previous esimates for species richness in the Gulf of Mexico for marine invertebrates were of `r format(11150-1541-434, big.mark=",")` species, and `r format(8374-909-134)` for the Pacific coast (including California) [@Fautin2010].

@@

Another potential venue to improve the quality of the data in biodiversity information systems, is better incitives for scientisits to publish their occurrence records alongside their papers as it is done for genetic data with GenBank [@Huang2011].

@@

Assembling species lists is a necessary but not sufficient step to understand biodiversity. Beyond counting the number of species, we need to understand better the geographical distribution of species, their abundance, and their biology. In order to predict the impact of climate change and other anthropogenic perturbations on species distributions, we need to collect more occurences for species. One potential challenge for this is to be able to convey taxonomic concepts for mostly unamed species. Developing robust identifiers that can be used across biodiversity information systems, and potentially anchored in results from DNA barcoding or other molecular techniques, is needed to collaboratively build our knowledge of species distributions.

@@

Given the urgency of the biodiversity crisis, we need to develop data-driven initiatives to efficiently document biodiversity, and build on our existing knowledge to collect information to improve our understanding of marine systems.
Digitization efforts are catching up on an already fragmented and vastly incomplete records.
